\section{Problem Statement, Overview, and Objectives}

%\subsection{Problem Statement}
%\input{problem}

Differential testing has long been considered one of the most powerful
approaches to ensure the security, reliability, and correctness of
software, \emph{when applicable}.  Differential testing offers a unique
solution to the difficulty of specifying what a software system should
(and should not) do: in essence, it tests a function $f$ by checking
its behavior matches function $g$ that is \emph{an independent
  implementation of the same functionality}, and defines test failures
as cases where $f(x) \neq g(x)$.  The ``functions'' $f$ and $g$ in
realistic settings are seldom actual code-level functions, of course,
but more typically complex systems ranging from data-structure
implementations to complex parsers and, especially, compilers and file
systems.

The ubiquity of differential approaches to testing compilers and file
systems is no accident:  these are two critical pieces of system
infrastructure where genuinely independent implementations are often
available.  Of course, since file systems may rest on different
hardware substrates, and compilers may make differenct choices where a
language allows for implementation defined behavior, in practice even
in these best-case scenarios, the problem is usually not one of checking that $f(x)
= g(x)$, but of checking that $O_f(f(I_f(x))) = O_g(g(I_g(x)))$, where
$O_f$, $O_g$, $I_f$, and $I_g$ are respectively, output and input
abstraction functions for $f$ and $g$, that allow for known
differences in behavior and input spaces.

Despite these complexities, differential testing is extremely
attractive, especially in settings where inputs are automatically (and
stochastically) generated, such as random testing and \emph{fuzzing}.
Indeed, arguably the fundamental limitation of modern fuzzing, despite
its huge success in finding software vulnerabilities and bugs, is the
typical weakness of fuzzing oracles.  It is extremely hard to
determine if a system is behaving correctly on an arbitrary, often
extremely exotic, input.  This leads to results such as the difference
between functional tests and fuzzing harnesses for Bitcoin core, where
equivalent, almost 100\% branch and code coverage yields huge
discrepancies in mutation detection, with functional tests detecting
over 80\% of injected bugs while fuzzing, despite perhaps more
aggressive exploration of the state space, detects only 12\% of these
bugs.  A differential specification makes it possible for fuzzing to
detect subtle semantic errors, not just memory safety violations and
assertion failures.

It seems likely that the importance of differential fuzzing will only
grow in the future, for two reasons, both deriving from the use of
LLMs and agentic coding.  First, the primary reason differential
fuzzing is seldom used in practice is that the cost of an independent
implementation of a software system, even in security or safety
critical settings, is prohibitive.  This cost is especially high given
that previous work has shown that ``independent'' implementations are
seldom as independent (in terms of correlation of bugs) as desirable.
Only in cases of extremely complex specifications implemented by
radically separated teams (e.g., compilers and file systems) is this
cost manageable, in large part because in these unusual settings a
second implementation is essentially free.  Code synthesis by AI
models sidesteps this problem:  once a specification that produces a
single workable implementation is available, having a single model
produce multiple implementations is often relatively cheap.  Using
multiple models or changing prompts to focus code generation on, for
example, a faster implementation, an implementation with a smaller
memory footprint, a more readable implementation, an implementation
using an unusual algorithm, and so forth, seems like to produce
controlled divergence that may exceed that of human developers.  If
the development cost of multiple implementations is radically reduced,
the advantages of differential specification are clear.

Second, differential approaches seem ideal for probing ambiguities in
specifications presented to AIs, and AI misunderstandings:  if
multiple implementations are produced from a specification, and
checked for equivalence, supervisors of AI-driven development can
examine cases where the AI's ``view'' of the software's behavior is
unstable, and can work to remove ambiguity and correct mistakes.

Differential fuzzing is, of course, already an important practice,
particularly in security-critical code development.  However,
differential fuzzing as practiced now is often limited in
effectiveness.  All too often, the approach taken is to independently
fuzz two implementations, then compare their behavior over the union
of the produced corpuses of interesting inputs.  While this can
discover bugs, it is effectively throwing away most of the power of
fuzzing, in two important ways.  First, the implementations are only
tested differentially for a small number of inputs, the final results
of fuzzing runs on the two separate systems; this missed out on the
opportunity to detect divergence and bugs on potentially millions or
billions of executions.  Second, the two independent corpuses are only
derived from efforts to cover behaviors of the two systems
independently; is an inputs does not produce novelty in terms of either system
in isolation, but represents a unique \emph{pair of coverage patterns} for a
shared input, this method is simply incapable of identifying and
exploiting such inputs.  Such inputs, however, are, intuitively, the
most promising in terms of detecting divergences in implementation behaviors.

With more effort, users of differential fuzzing can compile two
implementations in such a way that a fuzzer can see the combined
coverage of both systems, essentially a program path of {\tt r1 =
  f(x); r2 = g(x); assert (r1==r2);}.  This overcomes the above
objections, but still in fact leaves much to be desired, in terms of
taking advantage of the differential nature of the fuzzing problem.

\section{Intellectual Merit}
\input{merit}